---
title: "Class 08: Breast Cancer Analysis Project"
author: "Barry (PID: 911)"
format: pdf
toc: true
---

## Background

The goal of todays mini-project is to explore a complete analysis using the unsupervised learning techniques covered in the last class. We will extend what we learned by combining PCA as a preprocessing step to clustering using data that consist of measurements of cell nuclei of human breast masses.

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

Values in this data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass.


## Data import

The data is available as a CSV from the class website:

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
```

Make sure we do not include sample id or diagnosis columns in the data that we analyze below.

```{r}
diagnosis <- as.factor(wisc.df$diagnosis)
wisc.data <- wisc.df[, -1]
dim(wisc.data)
```

```{r}
head(wisc.data)
```

## Exploratory data analysis

> Q1. How many observations are in this dataset?

There are `r nrow(wisc.data)` observations/samples/patients in the data set.


> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(wisc.df$diagnosis == "M")
```
```{r}
table(wisc.df$diagnosis)
```


> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
#colnames(wisc.data)
length( grep("_mean", colnames(wisc.data)) )
```
```{r}
n <- colnames(wisc.data)
inds <- grep("_mean", n)
length(inds)
```

## Principal Component Analysis

The main function in base R for PCA is called `prcomp()`. A optional argument `scale` should nearly always be switched to `scale=TRUE` for this function.

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

Let's make our main result figure - the "PC plot" or "score plot", "ordienation plot"...


```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```


> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

```{r}
pr.var <- wisc.pr$sdev^2
pr.var / sum(pr.var) 
```


> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?


## Hierarchical clustering

```{r}
apply(scale(wisc.data), 2, sd)
```



```{r}
d <- dist(scale(wisc.data))
h <- hclust(d)
plot(h)
```

# Combining PCA and clustering

```{r}
d <- dist( wisc.pr$x[,1:3] )
wisc.pr.hclust <- hclust(d, method="ward.D2")
plot(wisc.pr.hclust)
abline(h=70, col="red")
```

Get my cluster membership vector
```{r}
grps <- cutree(wisc.pr.hclust, h=70)
table(grps)
```
```{r}
table(diagnosis)
```
Make a wee "cross-table"

```{r}
table(grps, diagnosis)
```
TP: 179
FP: 24

Sensitivity: TP/(TP+FN)


