---
title: "Class05: Data Viz with GGPLOT"
author: "Barry (PID: 911)"
format: pdf
---

Today we playing with plotting and graphics in R.

There are lots of ways to make cool figures in R.
There is "base" R graphics (`plot()`, `hist()` `boxplot()` etc.) 

There is also add-on packages, like **ggplot**.

```{r}
head(cars, 3)
```
Let's plot this with "base" R:

```{r}
plot(cars)
```

```{r}
head(mtcars)
```

Let's plot `mpg` vs `disp`

```{r}
plot(mtcars$mpg, mtcars$disp, pch=10)
```
```{r}
hist(mtcars$mpg)
```

## GGPLOT

The main function in the ggplot2 package is `ggplot()`. First I need to install the **ggplot2** package. I can insall any packge with the function `install.packages()` 

> **N.B.** I never want to run `install.packgages()` in my quarto source document!!!


```{r}
library(ggplot2)
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```
```{r}
ggplot(cars) +
  aes(speed) +
  geom_histogram()
```

Every ggplot needs at least 3 things:

- The **data** (given with `ggplot(cars)`)
- The **aesthetic** mapping (given with `aes()`)
- The **geom** (given by `geom_point()`)

> For simple canned graphs "base" R is nearly always faster

### Adding more layers

Let's add a line and a tile, subtitle and caption as well as custom axis labels

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_line()
```

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
   geom_smooth(method="lm", se=FALSE) +
   labs(title = "Silly Plot", 
    x = "Speed (MPH)", 
    y = "Distance (ft)",
    caption = "when is tea time anyway???") +
  theme_bw()
```

## Plot some expresion data

Read data file from online URL

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q1. How many genes are in this wee dataset?

There are `r nrow(genes)` in this dataset

> Q2. How many "up" regulated genes are there?

```{r}
sum(genes$State == "up")
```
There are `r sum(genes$State == "up")` UP genes


```{r}
table(genes$State)
```

```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=Condition2) +
  geom_point()
```

```{r}
p <- ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point() +
  scale_color_manual(values=c("blue","gray","red"))
```

```{r}
p + labs(title="look at me")
```


Silly example of adding labels

```{r}
library(ggrepel)

ggplot(genes) +
  aes(x=Condition1, 
      y=Condition2, 
      col=State, 
      label=Gene) +
  geom_point() +
  scale_color_manual(values=c("blue","gray","red")) +
  geom_text_repel(max.overlaps = 60) +
  theme_bw()
```

## Going Further

Playing with some different layers and the gapmider dataset...

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder)
```

```{r}
tail(gapminder)
```

A first plot

```{r}
ggplot(gapminder) +
  aes(y=lifeExp, x=gdpPercap, col=continent, 
      size = pop) +
  geom_point(alpha=0.2) +
  facet_wrap(~continent) +
  theme_bw()
```


```{r}
p
```



```{r}
library(dplyr)


```
```{r}
library(patchwork)

# Setup some example plots 
p1 <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp))

p2 <- ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear))

p3 <- ggplot(mtcars) + 
  geom_smooth(aes(disp, qsec))

p4 <- ggplot(mtcars) + 
  geom_bar(aes(carb))

# Use patchwork to combine them here:
(p1 | p2 | p3) /
      p4
```

```{r}
(p1 | p2 | p3) / 
(p4)
```

