---
title: "Class 10: Halloween Project"
author: "Barry"
format: pdf
toc: true
---

## Background

It is Halloween today and to help get in the mood for tonight let's explore some candy preference data from 538.


## Data Import

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)
```


## Data Exploration

```{r}
flextable::flextable( head(candy) )
```

```{r, message=FALSE}
library(dplyr)

candy |> nrow()
```
```{r}
nrow(candy)
```

```{r}
round( mean(candy$winpercent), 4) 
```
```{r}
candy |> select(winpercent) 
```


```{r}
win <- candy$winpercent
win.mean <- mean(win)
round(win.mean)
```


```{r}
candy %>% select(winpercent)

candy |> 
  select(winpercent)
```
```{r}
colnames(candy)
```
> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins=20)
```

> Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy) +
  aes(winpercent) +
  geom_density()
```


> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
# 1. Find all chocolate candy in the dataset
choc.inds <- as.logical(candy$chocolate)
choc.candy <- candy[choc.inds,]

# 2. Extract their `winpercent` values
choc.win <- choc.candy$winpercent

# 3. Find the mean of these values
choc.mean <- mean(choc.win)

# 4-6. Do the same for fruity candy
fruit.win <- candy[candy$fruity==1,]$winpercent
fruit.mean <- mean(fruit.win)

# 7. Which mean value is higher
choc.mean
fruit.mean
```



> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```

> Q13. What are the five least liked candy types in this set?

```{r}
x <- c(10, 2, 5, 1)
order(x)
```

```{r}
ord.ind <- order(candy$winpercent)
head( candy[ord.ind, ], 5)
```


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail( candy[ord.ind, ], 5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, reorder( rownames(candy), winpercent)) +
  geom_col()
```

Add some color

```{r}
my_cols <- rep("black", nrow(candy))
my_cols[candy$chocolate==1] <- "chocolate"
my_cols[candy$bar==1] <- "brown"
my_cols[candy$fruity==1] <- "pink"
#my_cols

ggplot(candy) +
  aes(x=winpercent, 
      y=reorder( rownames(candy), winpercent)) +
  geom_col(fill=my_cols)
```

## Winpercent vs Pricepercent

```{r}
library(ggrepel)

my_cols[candy$fruity==1] <- "red"

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols)
```
## Correlation 


```{r}
cij <- cor(candy)
```

```{r}
library(corrplot)
corrplot(cij)
```

## Principal Component Analysis

The main function in base R for this `prcomp()` and we want to set `scale=TRUE` here:

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Let's look at our first main result figure - the "PC plot" or PC1 vs PC2

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols) +
  labs(title="Barry's candy space map")
```
Don't forget about your variable "laodings" - how the origional variables contribute to your new PCs...

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```




